---
title: "Lecture 4: Finding the MLE with iterated filtering"
author:
  - Qianying (Ruby) Lin
  - Jesse Wheeler
  - Spencer J. Fox
format:
  beamer:
    classoption: "t"
    link-citations: true
    keep_tex: true
    slide_level: 2
    section-titles: false
    aspectratio: 169
    include-in-header: "../_includes/header.tex"
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
    hyperrefoptions:
      - linktoc=all
  html:
    toc: false
  pdf:
    toc: false
editor_options:
  chunk_output_type: console
bibliography: ["../sbied.bib"]
editor:
  markdown:
    wrap: 72
---

```{r knitr_opts,include=FALSE,cache=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())
```

```{r prelims,echo=FALSE,cache=FALSE,purl=FALSE}
library(tidyverse)
library(pomp)
library(foreach)
library(doParallel)
library(doRNG)
options(stringsAsFactors=FALSE)
set.seed(1350254336)
```

# Where we are

## Objectives

By the end of this lecture you will be able to:

1.  Say what the **maximum-likelihood estimate** is, and why a standard
    optimizer cannot find it on a noisy ridge.
2.  Explain the idea behind iterated filtering, without pseudocode.
3.  Run `mif2` on your laptop and read its output.
4.  Say what each of the four knobs does, and what
    `cooling.fraction.50` actually means.
5.  Explain why the likelihood `mif2` reports is **not** your
    likelihood.
6.  Read a convergence plot the way an experienced modeller reads one.

# From scoring to maximizing

## The maximum-likelihood estimate

Before the break we built a machine that **scores** any fixed parameter
set: `pfilter` turns a parameter vector into a log-likelihood.

\vspace{2mm}

Now we want the **best** parameters, the ones that make the data most
likely: the **maximum-likelihood estimate**,
$$
\hat\theta = \argmax_{\theta}\, \loglik(\theta).
$$

\vspace{2mm}

The obvious move is to hand our scoring machine to an optimizer (`optim`,
Nelder-Mead) and let it climb. First: what does the surface look like?

## The surface is a knife-edge ridge

\begin{center}
  \includegraphics[width=0.9\textwidth]{figures/fig-likelihood-ridge.png}
\end{center}

\small The two **lines** on the left are the two **curves** on the right:
*along* the ridge (green, $R_0$ fixed) the fit barely changes; *across* it
(orange, $\gamma$ alone, so $R_0$ changes) it drops three to five times faster.

## The surface is a knife-edge ridge

-   This is not a hill with a summit. It is a long diagonal **ridge**.
-   The ridge runs along **constant $R_0 = \beta/\gamma$**.

\vfill

The epidemiology, which matters more than the geometry:

> These data pin down $R_0$ **sharply**. They barely constrain $\beta$
> and $\gamma$ **separately**.

\vfill

You can trade a faster transmission rate against a faster recovery rate
and the epidemic looks the same. The data cannot tell those apart, so
neither can any method, no matter how good.

\vfill

\textbf{So what?} $R_0$ is what these data actually measure. Report it with
confidence; report $\beta$ on its own and you are claiming precision that is
not there. You will meet this ridge again in a moment, when twenty `mif2`
searches agree on the fit but disagree about $\beta$ by 50\%.

## And the surface is noisy

Every evaluation is a *different* random estimate. The optimizer does not
see the smooth ridge. It sees this:

\begin{center}
  \includegraphics[width=0.9\textwidth]{figures/fig-nelder-mead-fails.png}
\end{center}

## Why the obvious thing fails

Standard optimizers assume the function is:

-   **cheap**: ours takes seconds per evaluation;
-   **smooth**: ours is jagged with Monte Carlo noise;
-   **deterministic**: ours gives a different answer every call.

\vfill

They are built to walk downhill on a smooth landscape. Hand them a noisy
one and they walk confidently into the noise and stop.

\vfill

\begin{center}
\textbf{The optimizer isn't broken. The function you handed it is.}\\
We need an algorithm built for this: iterated filtering.
\end{center}

# The idea

## Stop fighting the noise. Use it.

The particle filter already does something remarkable: it keeps a swarm
of guesses about the **hidden state**, kills the bad ones, and clones
the good ones.

\vspace{4mm}

Iterated filtering asks one question:

\vspace{2mm}

> *What if we let the **parameters** wander too, and let the same
> selection process do the work?*

\vspace{4mm}

-   Give every particle its own slightly different parameters.
-   Run a particle filter. Particles that fit the data survive.
-   Their parameters survive with them.
-   Repeat, with a little less wandering each time.

## Iterated filtering, in one picture

\begin{center}
  \includegraphics[height=7cm]{figures/mif_graph.pdf}
\end{center}

## `mif2` is `pfilter` plus two boxes

::::: columns
::: {.column width="50%"}
\begin{center}
  \includegraphics[height=5cm]{../graphics/lec3pf.png}
\end{center}
\begin{center}\small before the break: \texttt{pfilter}\end{center}
:::

::: {.column width="50%"}
\begin{center}
  \includegraphics[height=5cm]{../graphics/lec4 pf.png}
\end{center}
\begin{center}\small now: \texttt{mif2}\end{center}
:::
:::::

\vfill

Same loop. Two new blue boxes: **perturb the parameters**, and
**resample them along with the states**. Plus an outer loop.

# Running it

## The four knobs

```{r mif2-demo,eval=FALSE}
po |> mif2(
  Np=2000,                     # particles
  Nmif=50,                     # iterations
  cooling.fraction.50=0.5,     # how fast perturbations shrink
  rw.sd=rw_sd(                 # how big perturbations start
    Beta=0.02, Rho=0.02, Eta=ivp(0.02)
  )
) -> mf

plot(mf)
```

\vfill

-   `Np`, `Nmif`: how much compute you are willing to spend.
-   `rw.sd`: how far parameters wander per step.
-   `ivp()`: marks a parameter that only affects the *start* of the
    epidemic, so it is only perturbed there.

## `cooling.fraction.50` is a promise about iteration 50

\begin{center}
  \includegraphics[width=\textwidth]{figures/fig-if2-cooling.png}
\end{center}

## `cooling.fraction.50` is a promise about iteration 50

`cooling.fraction.50=0.5` means:

> *After **fifty** iterations, the perturbations are **half** the size
> they started at.*

\vfill

It does **not** mean "halve them every iteration", and it does not mean
"halve them every fifty iterations". The decay is smooth and continuous,
and it keeps going after iteration 50.

\vfill

-   Cool too fast (0.1): the search freezes wherever it happens to be.
-   Cool too slow (0.9): the search never settles.

\vfill

\begin{center}
\textbf{Exercises 1 and 2.} \quad Open \texttt{4\_mle\_mif2.Rproj}, then
\texttt{scripts/EXERCISES.R}.
\end{center}

## One search tells you almost nothing

Run twenty. They are independent, so run them in parallel.

```{r mif2-parallel,eval=FALSE}
measSIR |> pomp(       # build the transform ONCE, outside the loop
  partrans=parameter_trans(log="Beta",logit=c("Rho","Eta")),
  paramnames=c("Beta","Rho","Eta")
) -> po

foreach(i=1:20, .combine=c, .packages="pomp", .export="po",
        .options.RNG=482947940) %dorng% {
    po |> mif2(Np=2000, Nmif=50, cooling.fraction.50=0.5,
             rw.sd=rw_sd(Beta=0.02, Rho=0.02, Eta=ivp(0.02))) -> mf
    mf
} -> mifs_local
```

\vfill

Build `parameter_trans` **outside** the loop. Inside, it forces a
recompile on every worker, a classic source of grief.

# Reading the output

## Twenty searches at once

\begin{center}
  \includegraphics[width=.9\textwidth]{figures/fig-mif2-traces.png}
\end{center}

## The most important plot in this module

Look at those two facts together:

-   The **log-likelihood** panel converges. All twenty searches agree
    the fit is good.
-   The **parameter** panels do not. $\beta$ disagrees by 50%.

\vfill

Both are true at once, and **neither is a bug**.

\vfill

The twenty answers are not a scattered cloud. They lie almost exactly on
a **line**: higher transmission, spread through a smaller pool of
susceptibles, with more of it reported, produces very nearly the same 42
numbers.

\vfill

\begin{center}
This is that same knife-edge ridge, now seen from the inside.
\end{center}

## Two different questions

::::: columns
::: {.column width="50%"}
**"Is my fit good?"**

Read the log-likelihood panel.

Here: yes. All twenty agree.
:::

::: {.column width="50%"}
**"Do I know this parameter?"**

Read that parameter's panel.

Here: **no**. Not $\beta$, not on its own.
:::
:::::

\vfill

A convergence plot answers both questions, and beginners routinely read
only the first one.

\vfill

\begin{center}
\textbf{Exercise 3.} \quad Run twenty searches and read the traces.
\end{center}

# When a search does not converge

## Still climbing? Run it longer

The healthy trace flattens. If your log-likelihood panel is **still rising**
at the last iteration, the search has simply not finished.

\vfill

`mif2` can be **continued**: hand a finished run back to `mif2` and it re-runs
from your current best parameters.

```{r continue-demo,eval=FALSE}
# within a foreach loop, or on a single run, we previously had:
po |> mif2(Np=2000, Nmif=50, cooling.fraction.50=0.5,
             rw.sd=rw_sd(Beta=0.02, Rho=0.02, Eta=ivp(0.02))) -> mf
# add this line to continue it with 50 more iterations:
mf |> mif2(Nmif=50) -> mf
```

\vfill

-   Continue each of the 20 runs, then re-plot the traces.
-   Rule of thumb: if the loglik panel still trends upward at the end, add
    more iterations. (You met continuation in **Exercise 2**.)

# The number `mif2` prints is not your likelihood

## Is the reported likelihood the true likelihood? NO!

- Always re-filter the final parameter vector to get a **TRUE** likelihood.

\begin{center}
  \includegraphics[width=0.95\textwidth]{figures/fig-why-reestimate.png}
\end{center}

## Why the reported number lies

Two reasons stack up:

1.  During a `mif2` run the parameters are **deliberately perturbed**.
    The likelihood being reported belongs to a *noisier model than
    yours*.

2.  `mif2` records the likelihood of the parameters at the **start** of
    each iteration. So your final parameter vector (the one in
    `coef(mf)`, the one you keep) has **never been filtered at all**.

\vspace{2mm}

Look at `traces(mf)`: the last log-likelihood is `NA`. That is not a
bug. It is the point.

\vfill

```{r reestimate-demo,eval=FALSE}
evals <- replicate(10, logLik(pfilter(mf, Np=5000)))
logmeanexp(evals, se=TRUE)
```

\vfill

\begin{center}
\textbf{Exercise 4.} \quad Re-estimate honestly, and compare.
\end{center}

## Stalled or dropping? Diagnose, do not panic

A trace that **dips or wanders** is not automatically a failure. The number
`mif2` reports is the likelihood of the **perturbed** model, estimated by a
particle filter, so it is random and need not climb every step.

\vfill

If it genuinely will not improve, check the usual culprits:

-   **Too few particles** (`Np`): the estimate is noisy and biased low, and the
    filter may be losing particles. Raise `Np`, and check the effective sample
    size.
-   **`rw.sd` too large**, or **cooling too slow**: the swarm never settles.

\vfill

\begin{center}
\textbf{The verdict never comes from the trace.} Re-filter the final
parameters with \texttt{pfilter}, on your real, unperturbed model.
\end{center}

\vspace{1mm}

\small Reference: the \texttt{pomp} FAQ 5.2,
\url{https://kingaa.github.io/pomp/FAQ.html}

# Exercises

## The exercises at a glance

1.  **Your first `mif2`.** Run one search; see what it changed.
2.  **Turn the knobs.** Continue a run; feel what `cooling.fraction.50` does.
3.  **Many searches at once.** Run 20; read the log-likelihood panel and the
    parameter panels as *different questions*.
4.  **The reported number lies.** Re-filter each endpoint honestly, keep the
    best.
5.  **(stretch) Does the start matter?** Scatter the starting points. Do they
    all arrive at the same place?
6.  **(lunch break) Does the latent period earn its keep?** Fit the **SEIR**
    you built in Lecture 2 and compare it to your SIR. One extra parameter,
    worth it?

# Where this fits

## Why this algorithm, and not another

::::: columns
::: {.column width="52%"}
Many methods have been proposed for POMP models. Three questions sort
them out:

-   Does it need a **formula** for the process, or will simulation do?
-   Does it use **all** the data, or just summaries?
-   Does it need a **prior**?

\vspace{3mm}

Iterated filtering: simulation is enough, uses all the data, no prior
required.
:::

::: {.column width="48%"}
\begin{center}
  \includegraphics[width=\linewidth]{../graphics/lec4 table.png}
\end{center}
:::
:::::

\vfill

\small That combination is why we teach it first [@Ionides2015; @King2016].

## Where we've got to

-   Iterated filtering lets the **parameters** ride along in the
    particle filter, and lets selection improve them.
-   Perturb, filter, resample, repeat, **cooling** the perturbations
    as you go.
-   `cooling.fraction.50` is a promise about **iteration 50**.
-   One search tells you nothing. **Run many.**
-   The log-likelihood panel and the parameter panels answer
    **different questions**.
-   The number `mif2` reports is not your likelihood. **Always
    re-filter.**

\vfill

\begin{center}
\textbf{Next, a full workflow: diagnostics, global searches,\\
and how to budget the compute.}
\end{center}

# References

## References

\small

::: {#refs}
:::

## License, acknowledgments, and links

-   This lesson is prepared for the [Simulation-based Inference for
    Epidemiological Dynamics](https://rubbislam.quarto.pub/episim/)
    module at the Summer Institute in Statistics and Modeling in
    Infectious Diseases,
    [SISMID](https://sph.emory.edu/SISMID/index.html).

-   The materials build on [previous versions of this course and related
    courses](../acknowledge.html).

-   Licensed under the [Creative Commons Attribution-NonCommercial
    license](https://creativecommons.org/licenses/by-nc/4.0/). Please
    share and remix non-commercially, mentioning its origin.
    \includegraphics[height=12pt]{../graphics/cc-by-nc}

-   Produced with R version `r getRversion()` and pomp version
    `r packageVersion("pomp")`.

\vfill

[`R` code for this lesson](./scripts/EXERCISES.R)

# Supplement: the theory we skipped

## The maximum-likelihood estimate

For reference: the same ideas as the lecture's pictures, written as the
formulas from the original *iterated filtering* lesson. None of this is
needed for the exercises.

Lecture A learned to *evaluate* the likelihood at a fixed $\theta$,
$\lik(\theta)=f_{Y_{1:N}}(y^*_{1:N};\theta)$. This lecture *maximises* it.
The **maximum-likelihood estimate** is
$$
\hat\theta = \argmax_{\theta}\, \loglik(\theta),
\qquad
\loglik(\theta) = \log \lik(\theta).
$$

Two facts make this hard, and both shape the algorithm. Each value
$\loglik(\theta)$ is a **noisy, expensive** particle-filter estimate, and
the surface is a near-flat **ridge** along constant $\Rzero=\beta/\gamma$.
So IF2 never differentiates $\loglik$ and never trusts a single evaluation:
it is a stochastic-approximation search built for exactly this surface.

## Iterated filtering (IF2): inputs and knobs

**Input.** Simulators for the initial and transition densities
$f_{X_0}(x_0;\theta)$ and $f_{X_n\mid X_{n-1}}(x_n\mid x_{n-1};\theta)$; an
evaluator for the measurement density $f_{Y_n\mid X_n}(y_n\mid x_n;\theta)$;
and the data $y^*_{1:N}$. It calls `rprocess` and `dmeasure` only, never
`dprocess`: IF2 is a **plug-and-play** method.

\textbf{Algorithmic parameters} (with their \texttt{mif2} arguments):

-   `Nmif` $=M$: number of filtering iterations.
-   `Np` $=J$: number of particles.
-   initial parameter swarm $\{\Theta^{0}_{j}\}_{j=1}^{J}$.
-   `rw.sd`: per-parameter random-walk sd, squared into a diagonal
    variance matrix $V_n$.
-   `cooling.fraction.50` $=a$: how fast the perturbations shrink.

**Cooling.** At iteration $m$ the perturbation variance is
$V_n\,a^{2m/50}$, so the standard deviation carries $a^{m/50}$. Thus $a=0.5$
halves the sd by iteration $50$, quarters it by $100$, and so on: smooth,
continuous, and never quite zero. **Output:** the final swarm
$\{\Theta^{M}_{j}\}$, whose mean or median is $\hat\theta$.

## The IF2 algorithm {.allowframebreaks}

**Procedure.**

1.  For $m$ in $1{:}M$:
2.  $\quad$ *perturb:*
    $\Theta^{F,m}_{0,j}\sim \normal\big(\Theta^{m-1}_{j},\,V_0\, a^{2m/50}\big)$,
    $j=1{:}J$
3.  $\quad$ *initialise:*
    $X^{F,m}_{0,j}\sim f_{X_0}(x_0;\Theta^{F,m}_{0,j})$, $j=1{:}J$
4.  $\quad$ For $n$ in $1{:}N$:
5.  $\qquad$ *perturb:*
    $\Theta^{P,m}_{n,j}\sim \normal\big(\Theta^{F,m}_{n-1,j},\,V_n\, a^{2m/50}\big)$
6.  $\qquad$ *predict:*
    $X^{P,m}_{n,j}\sim f_{X_n\mid X_{n-1}}(x_n\mid X^{F,m}_{n-1,j};\Theta^{P,m}_{n,j})$
7.  $\qquad$ *weight:*
    $w^{m}_{n,j}= f_{Y_n\mid X_n}(y^*_n\mid X^{P,m}_{n,j};\Theta^{P,m}_{n,j})$
8.  $\qquad$ *resample:* draw $k_{1{:}J}$ with
    $\prob{k_j=i}= w^{m}_{n,i}\big/\sum_{u=1}^{J} w^{m}_{n,u}$
9.  $\qquad$ set $\Theta^{F,m}_{n,j}=\Theta^{P,m}_{n,k_j}$ and
    $X^{F,m}_{n,j}=X^{P,m}_{n,k_j}$
10. $\quad$ End For
11. $\quad$ Set $\Theta^{m}_{j}=\Theta^{F,m}_{N,j}$, $j=1{:}J$
12. End For

\framebreak

**Reading the algorithm.**

-   The inner loop (lines 4–10) is exactly the particle filter of Lecture A,
    run on a model whose parameters get a fresh perturbation every step.
-   The outer loop (lines 1–12) repeats that filter $M$ times, shrinking the
    perturbations each pass. That shrinkage is the cooling.
-   Superscript $F$ marks the *filtering* distribution (particles given
    $y^*_{1:n}$); superscript $P$ marks the *prediction* distribution
    (given $y^*_{1:n-1}$).
-   The weight $w^{m}_{n,j}$ is the likelihood of observation $n$ under
    particle $j$: fitter particles are resampled more often, and their
    perturbed parameters ride along.
-   As $J\to\infty$ with a suitable cooling schedule, the swarm concentrates
    on $\hat\theta$ [@Ionides2015]. IF2 is the only current
    **full-information, plug-and-play, frequentist** method for POMP models.
