---
title: "`phylopomp`: an R package for POMP inference on genealogies"
author:
  - Spencer J. Fox
  - Qianying (Ruby) Lin
  - Jesse Wheeler
format: 
  beamer:
    classoption: "t"
    # fontsize: "10pt"
    link-citations: true
    keep_tex: true
    slide_level: 2
    section-titles: false
    aspectratio: 169
    include-in-header: "../_includes/header.tex"
    # beameroptions:
    #   - setbeamertemplate: "footline[frame number]"
    #   - setbeamertemplate: "navigation symbols{}"
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
    hyperrefoptions:
      - linktoc=all
  html:
    toc: false
  pdf:
    toc: false
editor_options: 
  chunk_output_type: console
bibliography: ["../sbied.bib"]
editor: 
  markdown: 
    wrap: 72
---

```{r knitr_opts,include=FALSE,cache=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())
```

```{r load-packages, echo=FALSE}
library(tidyverse)
library(pomp)
library(ggplot2)
library(reshape2)
```

## Background: genomic sequences \& phylogenetic tree
\begin{center}
  \includegraphics[height=7cm]{../graphics/nextstrain_geneal.png}
\end{center}

## Background: phylogenetics \& phylodynamics
\begin{center}
  \includegraphics[height=7cm]{../graphics/phylogen_phylodyn.pdf}
\end{center}

The goal of `phylopomp` is to infer the partially observed epidemiological dynamics, formulated by compartmental models, 
using the genealogy/phylogeny, reconstructed from sampled genomic sequences.

## Installation

The github repo for `phylopomp` is on \url{https://github.com/kingaa/phylopomp/}.

We can install the latest released version from github using `devtools`.

```{r install, eval=FALSE}
library(devtools)
install_github("kingaa/phylopomp@0.14.8.0")
```

To confirm the package is installed successfully,

```{r install-check}
packageVersion("phylopomp")
library(phylopomp)
```

## Simulation and inference
- The first useful function of `phylopomp` is to build customized models and simulate genealogies from it.
Pre-defined models are included: linear birth-death model (`lbdp`), moran model, SIR, SEIR, two-class SIR model with super-infection (`si2r`), two-strain SIR model (`siir`), etc.
- The second is to infer the models given genealogies using `pomp`.
This function is currently under development, 
while a few models are available: `lbdp`, `moran`, `sir`, and `seir`.

## Example: simulation, pre-defined model {.allowframebreaks}
```{r sim-sir, fig.width=6, fig.height=4, dpi=300}
#| out-height: 55%
set.seed(1234)
simulate("SIR",Beta=2,gamma=1,psi=2,S0=1000,I0=5,time=2) |>
  simulate(Beta=5,gamma=2,time=5,psi=3) -> model.sir # update params
plot(model.sir)
```

\framebreak

```{r sim-seir1, fig.width=6, fig.height=4, dpi=300}
#| out-height: 55%
set.seed(1234)
simulate("SEIR",Beta=2,sigma=2,gamma=1,psi=2,S0=1000,I0=5,time=2) |>
  simulate(Beta=5,gamma=2,time=5,psi=3) -> model.seir
plot(model.seir)
```

\framebreak

```{r sim-seir2, fig.width=6, fig.height=4, dpi=300}
#| out-height: 65%
plot(model.seir, obscure=FALSE)
```

\framebreak
```{r sim-seir3, fig.width=6, fig.height=4, dpi=300}
#| out-height: 70%
lineages(model.seir, obscure=FALSE) |> plot()
```

## Example: inference, pre-defined model {.allowframebreaks}
```{r infer-sir}
library(pomp)
set.seed(1234)
# simulate a geneal object from an SIR model
simulate("SIR",Beta=3,gamma=1,psi=2,omega=1,S0=100,I0=5,time=10) -> x
# build a pomp object
x |>
  sir_pomp(
    Beta=3,gamma=1,psi=2,omega=1,
    S0=100,I0=5,R0=0
  ) -> po
po |> pfilter(Np=5000) -> pf

pf |> logLik()
```

\framebreak

```{r infer-seir}
set.seed(1234)
simulate("SEIRS",      # simulate a genealogy from an SEIR model
  Beta=4,sigma=1,gamma=1,psi=1,omega=1,
  S0=100,E0=3,I0=5,R0=100, time=5
) -> G
G |>
  seirs_pomp(
    Beta=4,sigma=1,gamma=1,psi=1,omega=1,
    S0=100,E0=3,I0=5,R0=100
  ) |> pfilter(Np=1000) |> 
  replicate(n=20) |> concat() -> pf
pf |> logLik() |> logmeanexp(se=TRUE)
```


## License, acknowledgments, and links

-   This lesson is prepared for the [Simulation-based Inference for Epidemiological Dynamics](https://rubbislam.quarto.pub/episim/) module at the Summer Institute in Statistics and Modeling in Infectious Diseases, [SISMID](https://sph.emory.edu/SISMID/index.html).

-   The materials build on [previous versions of this course and related courses](../acknowledge.html).

-   Licensed under the [Creative Commons Attribution-NonCommercial license](https://creativecommons.org/licenses/by-nc/4.0/). Please share and remix non-commercially, mentioning its origin. \includegraphics[height=12pt]{../graphics/cc-by-nc}

-   Produced with R version `r getRversion()` and pomp version `r packageVersion("pomp")`.

-   Compiled on 2025-07-23.

\vfill

[Back to Lesson](index.html)

[`R` code for this lesson](./main.R)


