---
title: "Lecture 3: The likelihood of a POMP, and computing it with the particle filter"
author:
  - Qianying (Ruby) Lin
  - Jesse Wheeler
  - Spencer J. Fox
format:
  beamer:
    classoption: "t"
    link-citations: true
    keep_tex: true
    slide_level: 2
    section-titles: false
    aspectratio: 169
    include-in-header: "../_includes/header.tex"
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
    hyperrefoptions:
      - linktoc=all
  html:
    toc: false
  pdf:
    toc: false
editor_options:
  chunk_output_type: console
bibliography: ["../sbied.bib"]
editor:
  markdown:
    wrap: 72
---

```{r knitr_opts,include=FALSE,cache=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())
```

```{r prelims,echo=FALSE,cache=FALSE,purl=FALSE}
library(tidyverse)
library(pomp)
library(foreach)
library(doParallel)
library(doRNG)
options(stringsAsFactors=FALSE)
set.seed(1350254336)
```

# Where we are

## Objectives

By the end of this lecture you will be able to:

1.  Say what the likelihood of a POMP model *is*, in plain language.
2.  Explain why we cannot just simulate our way to it.
3.  Run `pfilter` on your own laptop and get a log-likelihood for a fixed
    set of parameters.
4.  Explain why that number is **random**, and what to do about it.
5.  Read the filter's diagnostics, and know when you have enough particles.

\vfill

This lecture is deliberately light on formulas. The ideas here are
geometric, and we are going to look at them rather than derive them.

## The shape of the whole module

::::: columns
::: {.column width="52%"}
-   We have a **model** and we have **data**.
-   We need a **method** that connects them.
-   Everything we want (parameter estimates, uncertainty, forecasts,
    model comparison) comes out the other side.

\vspace{3mm}

This session is about the method. Specifically: about **scoring** a set
of parameters.
:::

::: {.column width="48%"}
\begin{center}
  \includegraphics[height=4.5cm]{../graphics/lec3_overview}
\end{center}
:::
:::::

## The question, in one picture

\begin{center}
  \includegraphics[height=5.2cm]{figures/likely_graph.pdf}
\end{center}

\vfill

Two models. One dataset. **Which model is doing a better job?**

Each shaded band is a model **simulated many times**. We turn that question
into a *number*, the **likelihood**, and for stochastic models like these
the only way to compute it is by **simulating**.

## The likelihood, in one sentence

> **The likelihood is the probability that this model, with these
> parameters, would have produced exactly the data we actually saw.**

\vfill

-   Bigger is better.
-   A single likelihood means nothing on its own. It only means
    something **compared to another one**.
-   We work with the **log**-likelihood, because the numbers involved
    are astronomically small.

\vfill

That is the whole definition. Everything from here is about how to
*compute* it.

# What we see, and what we don't

## The problem: most of the epidemic is invisible

\begin{center}
  \includegraphics[height=5.5cm]{../graphics/lec3ll.png}
\end{center}

We observe the top row. We never see the bottom row: who is
susceptible, who is infected.

## The problem: most of the epidemic is invisible

\begin{center}
  \includegraphics[width=.9\textwidth]{figures/fig-data-vs-hidden.png}
\end{center}

\vfill

We have 42 numbers. Underneath them is an entire epidemic we never
observed, and *many different* hidden epidemics are consistent with
the same 42 numbers.

## So the likelihood is a hard sum

To score one parameter set honestly, we would have to ask:

\vspace{2mm}

> *Across **every possible** hidden epidemic, weighted by how likely
> each one is: how plausible are my 42 numbers?*

\vspace{4mm}

-   That is a sum over an enormous space of possibilities.
-   For our measles model there is **no formula** for it.
-   But notice: we can **simulate** hidden epidemics all day long.

\vfill

\begin{center}
\textbf{Simulating forward is easy. Scoring backward is hard.}\\
That asymmetry is the reason this module exists.
\end{center}

# Attempt 1: just simulate a lot

## The obvious idea

If we can simulate epidemics, why not simulate a *lot* of them and see
which ones look like the data?

\begin{center}
  \includegraphics[height=4.2cm]{figures/directsim_graph.pdf}
\end{center}

Simulate many trajectories, score each one against the data, average.
This is a perfectly correct method. It is also useless.

## Why it fails

\begin{center}
  \includegraphics[width=\textwidth]{figures/fig-direct-sim-fails.png}
\end{center}

## Why it fails

-   A simulation that drifts away from the data **never comes back**.
    Nothing pulls it home.
-   Over 42 weeks, essentially every trajectory drifts away.
-   So almost all the simulations contribute nothing, and the answer is
    decided by a **handful of lucky ones**.

\vfill

The method is not *wrong*. It is just spending all its effort on
epidemics that stopped resembling the data weeks ago.

\vfill

\begin{center}
\textbf{We need to stop bad trajectories early, not after 42 weeks.}
\end{center}

# The particle filter

## The fix: don't let bad guesses run

The particle filter changes one thing. Instead of letting a simulation
run to the end, it checks in **at every observation**:

\vspace{2mm}

1.  **Predict**: let every particle take one step forward.
2.  **Weight**: the data arrive. Particles that predicted well score
    highly.
3.  **Resample**: kill the bad ones, clone the good ones.

\vspace{2mm}

Then repeat, 42 times.

\vfill

\begin{center}
  \includegraphics[width=0.92\textwidth]{figures/pfilter_graph.pdf}
\end{center}

## One week of the particle filter, on real data

\begin{center}
  \includegraphics[width=.8\textwidth]{figures/fig-pfilter-one-week.png}
\end{center}

\vfill

Direct simulation lets bad trajectories run for 42 weeks. The particle
filter kills them **every single week** and clones the survivors.

## What the algorithm needs from you

::::: columns
::: {.column width="45%"}
The particle filter only ever asks your model two things:

-   *"Take a step forward."* (`rprocess`)
-   *"How well does this state explain this week's data?"*
    (`dmeasure`)

\vspace{3mm}

It never asks for a formula. That is why it works on models we cannot
write down.
:::

::: {.column width="55%"}
\begin{center}
  \includegraphics[height=5.2cm]{../graphics/lec3pf.png}
\end{center}
:::
:::::

## Doing it in `pomp`

```{r pfilter-demo,eval=FALSE}
source("scripts/model_measSIR.R")

pf <- measSIR |> pfilter(Np=1000)
logLik(pf)
```

\vfill

That is the whole thing. `Np` is the number of particles.

\vfill

\begin{center}
\textbf{Exercises 1 and 2.} \quad Open \texttt{3\_likelihood\_pfilter.Rproj}, then
\texttt{scripts/EXERCISES.R}.
\end{center}

\vspace{2mm}

Look at the model against the data, then compute your first
log-likelihood.

# The answer is a random number

## Run it twice, get two answers

The particle filter is a **Monte Carlo** algorithm. It does not return
the likelihood. It returns a *random estimate* of it.

\vspace{3mm}

```{r rerun-demo,eval=FALSE}
replicate(5, logLik(measSIR |> pfilter(Np=1000)))
```

\vspace{3mm}

Five runs, five different numbers. So: how many particles do we need,
and how wrong are we if we skimp?

## Too few particles doesn't just add noise

\begin{center}
  \includegraphics[width=0.87\textwidth]{figures/fig-loglik-is-random.png}
\end{center}

\vfill

\begin{center}
\textbf{In practice:} escalate \texttt{Np} = 2000, then 5000, then 10000.
\end{center}

## Too few particles doesn't just add noise

-   More particles makes the cloud **tighter**; that much is
    expected.
-   More particles also moves the cloud **up**.

\vfill

With too few particles you do not get a noisy answer. You get a
**confidently wrong** one, biased low.

\vfill

Practical rule: if adding particles keeps raising your log-likelihood,
you did not have enough particles.

## Where does the noise come from? A quick diagnostic

`pomp` gives you `plot(pf)`. Its key panel is the **effective sample
size**: at each week, how many of your particles are really contributing?

\begin{center}
  \includegraphics[width=0.8\textwidth]{figures/fig-ess-collapse.png}
\end{center}

## Where does the noise come from? A quick diagnostic

-   Where ESS is high, hundreds or thousands of particles agree; that
    week's contribution to the likelihood is solid.
-   Where ESS **crashes**, the whole week rests on a few hundred
    particles. That is where your estimate is fragile, and where the
    run-to-run noise is born.
-   Here the filter struggles at the **early rise** and at the **crash to
    zero**, not at the big mid-epidemic jump, which it handles fine.

\vfill

This is a *filter* health check: it tells you whether you have enough
particles, and where. Using it to criticize the *model* comes later, in
the workflow.

\vfill

\begin{center}
\textbf{Exercises 3 and 4.} \quad See the number jump around, then read the
diagnostic that explains why.
\end{center}

## Combining repeated runs

Run several filters and combine them, but **not** with `mean()`:

```{r logmeanexp-demo,eval=FALSE}
pfs |> logLik() |> logmeanexp(se=TRUE)
```

\vspace{3mm}

Read the name backwards: `log(mean(exp(x)))`.

-   The particle filter is unbiased for the **likelihood**.
-   It is *not* unbiased for the **log**-likelihood.
-   So we average on the likelihood scale, then take the log.

\vspace{3mm}

`logmeanexp` does this without overflowing, and gives you a standard
error.

## Each run is independent, so parallelize

```{r parallel-demo,eval=FALSE}
library(foreach); library(doParallel); library(doRNG)

cl <- makePSOCKcluster(detectCores()-1)
registerDoParallel(cl)
registerDoRNG(1234)

foreach(i=1:20, .combine=c, .packages="pomp", .export="measSIR") %dopar% {
  measSIR |> pfilter(Np=5000)
} -> pfs

stopCluster(cl)
```

\vfill

`makePSOCKcluster` works on macOS, Linux **and** Windows.
`registerDoRNG` makes the parallel random numbers reproducible.

\vfill

\begin{center}
\textbf{Exercises 5 and 6.} \quad Parallel evaluation, and what accuracy
costs you.
\end{center}

# Exercises

## The exercises at a glance

Everything builds to one idea: **the likelihood is a number we can compute,
and it is random.**

\vspace{1mm}

1.  **Look before you compute.** Simulate 20 outbreaks. Do they resemble the
    data?
2.  **Your first likelihood.** One `pfilter`, one log-likelihood.
3.  **Run it again.** The same line returns a *different* number. Why?
4.  **Read the diagnostics.** Effective sample size week by week: where does
    the filter struggle?
5.  **Run many, in parallel.** Combine 20 estimates with `logmeanexp` and a
    standard error.
6.  **What accuracy costs.** Time `pfilter` against `Np`: the price of
    precision.
7.  **(stretch) A likelihood slice.** Vary `Eta`, watch the likelihood move:
    a first look at the surface.

# Where that leaves us

## Where we've got to

We can now **score any fixed set of parameters**:

-   the particle filter turns a parameter vector into a log-likelihood;
-   it only asks the model to simulate and to weigh, so it works even on
    models we cannot write down;
-   the answer is random, so we repeat and combine with `logmeanexp`;
-   more particles buys accuracy, not just precision.

\vfill

But a score for *one* parameter set is not the answer. We want the set that
fits the data **best**: the maximum-likelihood estimate.

\vfill

\begin{center}
\textbf{After the break: which parameters maximize the likelihood, and the
algorithm that finds them.}
\end{center}

## If you want to go deeper

You do not need any of this to do the exercises. It is here for the people
who want the theory we skipped.

\vspace{3mm}

-   **The `pomp` package**, and the framework behind everything in this
    module: @King2016.
-   **The particle filter**: @Arulampalam2002 is the standard tutorial;
    @Kitagawa1987 is the original; @Doucet2001 is the reference volume.
-   **The likelihood**, as a general tool for inference: @Pawitan2001.
-   **Iterated filtering**, which is the next session: @Ionides2015.

\vfill

\begin{center}
The maths we skipped is all in these, and the key formulas are in the
supplement at the end of this deck.
\end{center}

# References

## References {.allowframebreaks}

\small

::: {#refs}
:::

## License, acknowledgments, and links

-   This lesson is prepared for the [Simulation-based Inference for
    Epidemiological Dynamics](https://rubbislam.quarto.pub/episim/)
    module at the Summer Institute in Statistics and Modeling in
    Infectious Diseases,
    [SISMID](https://sph.emory.edu/SISMID/index.html).

-   The materials build on [previous versions of this course and related
    courses](../acknowledge.html).

-   Licensed under the [Creative Commons Attribution-NonCommercial
    license](https://creativecommons.org/licenses/by-nc/4.0/). Please
    share and remix non-commercially, mentioning its origin.
    \includegraphics[height=12pt]{../graphics/cc-by-nc}

-   Produced with R version `r getRversion()` and pomp version
    `r packageVersion("pomp")`.

\vfill

[`R` code for this lesson](./scripts/EXERCISES.R)

# Supplement: the theory we skipped

## The likelihood of a POMP

For reference: the same ideas as the lecture's pictures, written as the
formulas from the original *likelihood* lesson. None of this is needed for
the exercises.

The data are $N$ observations $y^*_{1:N}$; the model is a density
$f_{Y_{1:N}}(y_{1:N};\theta)$. The **likelihood** evaluates it at the data,
and the **log-likelihood** is its log:
$$
\lik(\theta) = f_{Y_{1:N}}(y^*_{1:N};\theta),
\qquad
\loglik(\theta) = \log \lik(\theta).
$$

The Markov structure factors the joint density of states and observations
into an **initial**, a **transition**, and a **measurement** density:
$$
f_{X_{0:N},\,Y_{1:N}}(x_{0:N},y_{1:N};\theta)
= f_{X_0}(x_0;\theta)\!\prod_{n=1}^N\!
   f_{X_n\mid X_{n-1}}(x_n\mid x_{n-1};\theta)\,
   f_{Y_n\mid X_n}(y_n\mid x_n;\theta).
$$

## The likelihood is an integral, and simulating it directly fails

The likelihood marginalizes over **every possible hidden path** $x_{0:N}$, a
high-dimensional integral with no closed form:
$$
\lik(\theta) = \int f_{X_0}(x_0;\theta)\!\prod_{n=1}^{N}\!
   f_{Y_n\mid X_n}(y^*_n\mid x_n;\theta)\,
   f_{X_n\mid X_{n-1}}(x_n\mid x_{n-1};\theta)\, dx_{0:N}.
$$

Write it as an **expectation** over unconditional simulations
$X_{0:N}\sim f_{X_{0:N}}$, then approximate by an average over $J$ of them:
$$
\lik(\theta)
= \expect{\prod_{n=1}^{N} f_{Y_n\mid X_n}(y^*_n\mid X_n;\theta)}
\approx \frac{1}{J}\sum_{j=1}^{J}\prod_{n=1}^{N}
        f_{Y_n\mid X_n}(y^*_n\mid X^j_n;\theta).
$$

Correct, but simulated paths drift off the data, so the effort needed to
keep any near it grows explosively with $N$. That is why we filter instead.

## The particle filter {.allowframebreaks}

Factor the likelihood **one observation at a time**:
$$
\lik(\theta)=\prod_{n=1}^N \int
   f_{Y_n\mid X_n}(y^*_n\mid x_n;\theta)\,
   f_{X_n\mid Y_{1:n-1}}(x_n\mid y^*_{1:n-1};\theta)\, dx_n.
$$

Two steps carry the particle cloud from one observation to the next.
**Prediction** pushes it forward through the dynamics; **filtering**
reweights it by the new observation, via Bayes:
$$
f_{X_n\mid Y_{1:n-1}}
= \int f_{X_n\mid X_{n-1}}\, f_{X_{n-1}\mid Y_{1:n-1}}\, dx_{n-1},
$$
$$
f_{X_n\mid Y_{1:n}}
= \frac{f_{Y_n\mid X_n}(y^*_n\mid x_n;\theta)\; f_{X_n\mid Y_{1:n-1}}}
       {\displaystyle\int f_{Y_n\mid X_n}(y^*_n\mid u_n;\theta)\;
        f_{X_n\mid Y_{1:n-1}}\, du_n}.
$$

The filter estimates each factor by Monte Carlo, so the log-likelihood is a
sum of per-observation pieces:
$$
\loglik(\theta) = \sum_{n=1}^N \log \lik_n(\theta)
  \approx \sum_{n=1}^N \log \hat{\lik}_n(\theta).
$$

The estimate $\hat{\lik}(\theta)$ is **unbiased** for $\lik(\theta)$; taking
logs makes $\log\hat{\lik}(\theta)$ a consistent but **downward-biased**
estimate of $\loglik(\theta)$, which is why too few particles read low.
