---
title: "Lecture 10: Genealogies and phylodynamics with phylopomp"
author:
  - Qianying (Ruby) Lin
  - Spencer J. Fox
  - Jesse Wheeler
format:
  beamer:
    classoption: "t"
    link-citations: true
    keep_tex: true
    slide_level: 2
    section-titles: false
    aspectratio: 169
    include-in-header: "../_includes/header.tex"
    header-includes: |
       \setbeamertemplate{navigation symbols}{}
       \setbeamertemplate{footline}[page number]
    hyperrefoptions:
      - linktoc=all
  html:
    toc: false
  pdf:
    toc: false
editor_options:
  chunk_output_type: console
bibliography: ["../sbied.bib"]
editor:
  markdown:
    wrap: 72
---

```{r knitr_opts,include=FALSE,cache=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())
```

```{r prelims,echo=FALSE,cache=FALSE,purl=FALSE}
library(tidyverse)
options(stringsAsFactors=FALSE)
set.seed(1350254336)
```

# Where we are

## Objectives

So far the **data** were a time series of case counts. Now the data are a
**genealogy**: a family tree of the pathogen, reconstructed from sampled
genome sequences.

\vspace{2mm}

By the end of this lecture you will be able to:

1.  Say what a genealogy is, and how it links genome data to the
    transmission dynamics.
2.  Simulate a genealogy from a compartmental model with `phylopomp`.
3.  Read a tree and its **lineages-through-time**, and see the epidemic in
    their shape.
4.  Evaluate the **likelihood** of a model given a genealogy, using the
    *same* particle filter from Lectures A and B.

# From case counts to genealogies

## The pathogen carries its own history

As an epidemic spreads, the pathogen mutates from host to host. Sequencing
the samples, then reconstructing their phylogeny with tools built for it
(e.g. BEAST), yields a **genealogy**: an estimate of who infected whom, as a
tree. `phylopomp` begins where that pipeline ends, taking the tree as its
data.

\vspace{3mm}

\begin{center}
  \includegraphics[width=.7\linewidth]{../graphics/nextstrain_geneal.png}
\end{center}

## Phylogenetics asks *what*; phylodynamics asks *why*

\begin{center}
  \includegraphics[width=0.92\linewidth]{../graphics/phylogen_phylodyn.pdf}
\end{center}

`phylopomp` treats the genealogy as a **partial observation** of an
epidemic, and infers the transmission dynamics behind it, the same
POMP idea as the rest of the module: a hidden process, observed noisily.

## The tree is a partial observation

::::: columns
::: {.column width="60%"}
\begin{center}
  \includegraphics[width=.8\linewidth]{../graphics/MGP_events.pdf}
\end{center}
:::

::: {.column width="40%"}
\small

Births (transmissions), deaths (recoveries), and samples are events in the
hidden epidemic. We only ever see **some** of them, and the tree is what
survives. `phylopomp` takes that genealogy as given, observed exactly; here
we **simulate** trees, so the workflow is self-contained. The theory is in
@King2025.
:::
:::::

# The package

## Installing `phylopomp`

`phylopomp` lives on GitHub at
\url{https://github.com/kingaa/phylopomp}. Install the latest version,
then check it loaded:

```{r install,eval=FALSE,echo=TRUE}
# install.packages("pak")         # if you do not have it
pak::pak("kingaa/phylopomp")

library(phylopomp)
packageVersion("phylopomp")       # 0.19.4.2 or newer
```

\vfill

\begin{center}
\small It compiles C++ under the hood, so give it a minute the first time.
\end{center}

## Two things it does

`phylopomp` provides two families of tools.

\vspace{2mm}

-   **Simulate** a genealogy from a compartmental model, with one function
    per model: `runSIR`, `runSEIRS`, `runSIIR` (two strains), `runSI2R`
    (super-spreading), `runLBDP` (linear birth-death), and more.
-   **Infer**: turn a genealogy into a `pomp` object with `sir_pomp`,
    `seirs_pomp`, `si2rs_pomp`, `lbdp_pomp`, ..., and then use everything you
    already know, `pfilter`, `mif2`, `logmeanexp`.

\vfill

\begin{center}
\textbf{The inference tools are newer and still growing; a handful of
models are wired up so far.}
\end{center}

# Simulating a genealogy

## One function builds and simulates the tree

```{r sim-sir,eval=FALSE,echo=TRUE}
library(phylopomp)
set.seed(2024)
runSIR(
  time = 4,               # observation window, t = 0 .. 4
  Beta = 3, gamma = 1,    # transmission and recovery rates
  psi  = 0.4,             # sampling rate (how many tips you get)
  pop  = 500,             # total host population
  S0 = 0.97, I0 = 0.03    # initial FRACTIONS of pop (not counts)
) -> G

plot(G, points = TRUE)    # the genealogy
```

\vfill

\small Note the modern interface: the initial states are **fractions** of
`pop`, and each model has its own `run...` function.

## A simulated SIR genealogy

\begin{center}
  \includegraphics[width=0.75\linewidth]{figures/fig-sir-genealogy.png}
\end{center}

Blue dots are **samples** (the tips), green dots are **branch points**
(shared ancestors), brown dots are the **roots**. Time runs left to right.

## The tree's shape is the epidemic

`lineages()` counts how many branches are alive at each moment. It rises as
the epidemic grows and falls as it burns out.

```{r ltt,eval=FALSE,echo=TRUE}
plot(lineages(G))
```

\begin{center}
  \includegraphics[width=0.78\linewidth]{figures/fig-sir-ltt.png}
\end{center}

## Change the parameters partway through

`continue...` picks up a finished run and keeps going with new parameters,
handy for a control measure that drops transmission mid-epidemic.

```{r continue,eval=FALSE,echo=TRUE}
runSIR(time = 2, Beta = 3, gamma = 1, psi = 0.5,
       pop = 500, S0 = 0.98, I0 = 0.02) |>
  continueSIR(time = 5, Beta = 6, gamma = 2) -> G2   # new rates from t = 2
plot(G2)
```

# A richer model

## Two strains in one tree

`runSIIR` simulates two competing strains. With `obscure = FALSE`,
`phylopomp` keeps the strain labels and colours each lineage by the strain
it belongs to.

```{r siir,eval=FALSE,echo=TRUE}
set.seed(7)
runSIIR(
  time = 3, Beta1 = 4, Beta2 = 8,   # strain 2 spreads faster
  gamma = 1.5, psi1 = 0.6, psi2 = 0.6,
  pop = 300, S_0 = 0.95, I1_0 = 0.03, I2_0 = 0.02
) -> GI
plot(GI, obscure = FALSE)
```

## Two strains in one tree

\begin{center}
  \includegraphics[width=0.72\linewidth]{figures/fig-siir-strains.png}
\end{center}

Strain 2 (olive) spreads faster, so it seeds **more, larger** clades than
strain 1 (magenta). You can read the competition straight off the tree.

# Inference

## The likelihood of a genealogy

Inference is the same move as the rest of the module. Turn the tree into a
`pomp` object, then hand it to the **particle filter**:

```{r infer,eval=FALSE,echo=TRUE}
library(pomp)
G |>
  sir_pomp(Beta = 3, gamma = 1, psi = 0.4,
           S0 = 0.97, I0 = 0.03, R0 = 0, pop = 500) |>
  pfilter(Np = 1000) |>
  replicate(n = 8) |> concat() -> pf

logmeanexp(sapply(pf, logLik), se = TRUE) # same estimator as Lecture 3
```

\vfill

Nothing here is new. The **data** changed (a tree, not a case count), but
$\loglik(\theta)$, the particle filter, and `logmeanexp` are exactly as
before.

\vspace{2mm}

\small A real, dated tree enters the same pipeline via `parse_newick()`; we
simulate here to keep the lecture reproducible.

## The likelihood peaks at the truth

We simulated the tree at $\beta = 3$. Scoring a grid of $\beta$ (all else
fixed) with that particle filter puts the peak right where it should be.

\begin{center}
  \includegraphics[width=0.7\linewidth]{figures/fig-loglik-vs-beta.png}
\end{center}

From here, `mif2` (Lecture 4) maximises this likelihood exactly as it did
for the case-count model.

# Exercises

## Your turn

Open `10_phylopomp.Rproj`, then `scripts/EXERCISES.R`. Answers are in
`scripts/SOLUTIONS.R`.

\vspace{3mm}

1.  **Simulate and look.** Build an SIR genealogy, plot the tree and its
    lineages-through-time, then change the sampling rate `psi` and watch the
    tree fill in or thin out.
2.  **Two strains.** Simulate a two-strain `SIIR` genealogy and colour it by
    strain. Which strain leaves more descendants, and why?
3.  **Infer.** Fit the SIR model back to its own tree with `pfilter`, sweep
    `Beta`, and confirm the likelihood peaks near the truth.

\vfill

\begin{center}
\small A stretch exercise fits the \textbf{wrong} model to a tree and compares
it, by likelihood, with the model that made it.
\end{center}

# Where this fits

## Where we've got to

-   A **genealogy** is a partially observed epidemic: the same POMP idea,
    with a tree for data instead of a case-count series.
-   `phylopomp` **simulates** trees from compartmental models
    (`runSIR`, `runSIIR`, ...) and **turns them into `pomp` objects**
    (`sir_pomp`, ...).
-   Once it is a `pomp` object, **everything transfers**: `pfilter` scores
    it, `logmeanexp` de-biases, `mif2` maximises.
-   The tree's **shape**, its lineages-through-time, carries the dynamics
    that inference reads.

\vfill

\begin{center}
\textbf{Genome data, same pipeline. That is the payoff of the POMP
framework.}
\end{center}

# References

## References

\small

::: {#refs}
:::

## License, acknowledgments, and links

-   Produced with the `phylopomp` package of Aaron A. King and
    collaborators [@King2025].

-   This lesson is prepared for the [Simulation-based Inference for
    Epidemiological Dynamics](https://rubbislam.quarto.pub/episim/)
    module at the Summer Institute in Statistics and Modeling in
    Infectious Diseases,
    [SISMID](https://sph.emory.edu/SISMID/index.html).

-   The materials build on [previous versions of this course and related
    courses](../acknowledge.html).

-   Licensed under the [Creative Commons Attribution-NonCommercial
    license](https://creativecommons.org/licenses/by-nc/4.0/). Please
    share and remix non-commercially, mentioning its origin.
    \includegraphics[height=12pt]{../graphics/cc-by-nc}

-   Produced with R version `r getRversion()`, pomp version
    `r packageVersion("pomp")`, and phylopomp version
    `r packageVersion("phylopomp")`.

\vfill

[`R` code for this lesson](./scripts/EXERCISES.R)
