---
title: "Workflow part"
author:
  - Zian Zhuang
format: 
  beamer:
    keep_tex: true
    slide_level: 2
    section-titles: false
    aspectratio: 169
    beameroptions:
      - setbeamertemplate: "footline[frame number]"
      - setbeamertemplate: "navigation symbols{}"
  html:
    toc: false
  pdf:
    toc: false
editor_options: 
  chunk_output_type: console
bibliography: ["../sbied.bib"] 
---

# Workflow {.allowframebreaks}

```{r knitr_opts,include=FALSE,cache=FALSE,purl=FALSE}
source("../_includes/setup.R", local = knitr::knit_global())
```

```{r doFuture,echo=FALSE,cache=FALSE}
library(pomp)
library(doFuture)
library(microbenchmark)
plan(multisession)
set.seed(2488820)
```

data- initialization-

```{r meas-data1}
library(tidyverse)
read_csv(paste0("https://kingaa.github.io/sbied/stochsim/", 
  "Measles_Consett_1948.csv")) |> 
  select(week,reports=cases) -> meas
meas |> as.data.frame() |> head(n=3)
```

```{r meas-data2,echo=FALSE,fig.width=6, fig.height=3}
meas |>
  ggplot(aes(x=week,y=reports)) +
  geom_line() +
  geom_point()
```

```{r}
meas |> pomp(times="week",t0=0) -> measSIR
```

R、c-dmeasure rmeasure-

```{r}
sir_stoch <- Csnippet("
  double dN_SI = rbinom(S,1-exp(-Beta*I/N*dt));
  double dN_IR = rbinom(I,1-exp(-Gamma*dt));
  S -= dN_SI;
  I += dN_SI - dN_IR;
  R += dN_IR;
  H += dN_IR;
")
```

```{r more-csnips}
    sir_rinit <- Csnippet("
      S = nearbyint(Eta*N);
      I = 1;
      R = nearbyint((1-Eta)*N);
      H = 0;
    ")
    sir_dmeas <- Csnippet("
      lik = dnbinom_mu(reports,k,Rho*H,give_log);
    ")
    sir_rmeas <- Csnippet("
      reports = rnbinom_mu(k,Rho*H);
    ")
```

become pomp-

```{r sir_pomp}
measSIR  %>% 
  pomp(
    rprocess=euler(sir_stoch,delta.t=1/7),
    rinit=sir_rinit,
    rmeasure=sir_rmeas,
    dmeasure=sir_dmeas,
    accumvars="H",
    statenames=c("S","I","R","H"),
    paramnames=c("Beta","Gamma","N","Eta","Rho","k"),
    params=c(Beta=15,Gamma=0.5,Rho=0.5,k=10,Eta=0.06,N=38000)
  ) -> measSIR
```

simulation/likelihood with the initial guess

```{r, fig.width=8, fig.height=1.5}
measSIR |>
  simulate(nsim=20,format="data.frame",include.data=TRUE) |>
  ggplot(aes(x=week,y=reports,group=.id,color=.id=="data")) +
  geom_line() + guides(color="none")
```

check likelihood and ess

```{r}
pf <- measSIR |> pfilter(Np=5000)
plot(pf)
min(pf@eff.sample.size)
```

```{r}
foreach(i=1:10,.combine=c,
  .options.future=list(seed=TRUE)) %dofuture% {
  measSIR |> pfilter(Np=5000)
} -> pf
pf |> logLik() |> logmeanexp(se=TRUE) -> L_pf
L_pf

# Building up a picture of the likelihood surface II ?
pf[[1]] |> coef() |> bind_rows() |>
  bind_cols(loglik=L_pf[1],loglik.se=L_pf[2])
```



Effective Sample Size refers to the number of particles that effectively contribute to the approximation of the posterior distribution.


local maximum (IF2)

```{r}
foreach(i=1:20,.combine=c,
  .options.future=list(seed=482947940)
) %dofuture% {
  measSIR |>
    mif2(
      Np=2000, Nmif=50, cooling.fraction.50=0.5,
      rw.sd=rw_sd(Beta=0.02, Rho=0.02, Eta=ivp(0.02)),
      partrans=parameter_trans(
        log="Beta",logit=c("Rho","Eta")
      ),
      paramnames=c("Beta","Rho","Eta")
    )
} -> mifs_local
```

Iterated filtering diagnostics 

```{r}
mifs_local |>
  traces() |>
  melt() |>
  ggplot(aes(x=iteration,y=value,group=.L1,color=factor(.L1)))+
  geom_line()+
  guides(color="none")+
  facet_wrap(~name,scales="free_y")
```




based on the results, check the starting guess,

Then perform global maximum check likelihood get point est

do profile and obtain CI

check likelihood and ess

then prediction and compare with the observation
